Factors determining how catchments respond to forest cover change. Re-analysing global data sets.

### 2022-02-04 paper/Forest_and_Water/Forest_and_Water.Rmd
fixing page breaks and merge issues

---
title: Factors determining how catchments respond to forest cover change. Re-analysing global data sets.
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: DARE, The University of Sydney
    footnote: 1
  - name: Eliana Nervi
    email: eliananervi@gmail.com
    affiliation: INIA
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
    affiliation: IMFIA
address:
  - code: DARE
    address: ARC Training Centre Data Analytics for Resources and the Environment
  - code: The University of Sydney
    address: School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia
  - code: INIA
    address: Project Manager, FPTA 357, Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
  - code: IMFIA
    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow. However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation. The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow? Generalised additive models were used to run flexible regressions including multiple variables. 
  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow. Deforestation causes a 32\% greater change in flow compared to reforestation. Area of the catchment only affects the change in streamflow after log transformation, given the wide variety in the data from small scale paired watershed studies. Smaller studies dominate the database with 42\% of the data < 1 km2 and 65\% of the data < 10 km2. Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies. Despite these findings, overall explained variance (38\%) is low due the quality of the inputs and additional unknown confounding factors.
journal: "Journal of Hydrology"
date: "`r Sys.Date()`"
bibliography: forest_and_water.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
editor_options: 
  chunk_output_type: inline
---

@@

There has been an long and on-going discussion in the hydrologcal literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017]. The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa. The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover. However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchmensts after deforestation.

@@

In addition, additional variables added were the latitude and longitude for the centre of the catchment as an approximation of its spatial location. Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018]. For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.

@@

Several additional data points from catchment studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e. Table 6 in @zhang2011).
We also removed one data point from the analysis, which corresponds to catchment #1 (Amazon) in @zhang2017. This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded. Furthermore, the change in flow for catchment #76 was corrected from 600% to 157% after review of the original publication [@baker1984]. Finally, on review of all the data in @zhang2017 and @filoso2017 and the additional studies, 29 potential duplicates were identified and flagged in the data.

@@

## Statistical modelling

<!----
The starting point consists of the two tables in the supplementary data from Zhang et al. (2017), which means there is one table for small watersheds (< 1000 km2) and one table for large watersheds (> 1000 km2). We might combine these tables as we can easily add a variable indicating whether the watershed originates from the small or large table.
--->
```{r, echo = F, message = F}
require(tidyverse)
require(lubridate)
require(mgcv)
require(visreg)
require(patchwork)
suppressWarnings(require(knitr))
suppressWarnings(require(bookdown))
suppressWarnings(require(pander))

@@

<!-- read in the ET data -->
<!-- no longer needed as this is now included in the data
```{r, echo = F, warning = F, message=F}
PET_large <- read_csv("../../data/largeET.csv")
PET_small <- read_csv("../../data/smallET.csv")

@@

# 
PET_all <- bind_rows(PET_large, PET_small)
# There are 7 missing E0 values
# > PET_all$`Watershed #`[is.na(PET_all$E0)==T]
# [1]  38 278 269 243 236 299 250
```
--->

@@

```{r, eval = F, echo = F}
# checking missing values
summary(All_data)

<!--now combine with E0 values -->
<!-- No longer needed as now included in the data -->
```{r, echo = F}
# All_data <- left_join(All_data ,PET_all %>% select(-c(Longitude, Latitude)), by = "Watershed #")
# All_data <- All_data %>%
#   mutate(E0 = ifelse(is.na(E0.x)==T, E0.y, E0.x)) %>%
#   select(-c(E0.x,E0.y)) #%>%
#   #rename(Latitude = Latitude.x,
# #         Longitude = Longitude.x)

@@

Here $X_i$ are factorial variables, while $Z_i$ are continuous variables. The model assumes no direct interactions and all variables are additive. The changes in forest cover contain both positive (forestation) and negative values (deforestation). In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect if removing forest cover was the same as an equivalent reforestation. However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover. Therefore all the change in forest cover data is converted to positive values, and an additional column ($sign_{forest cover}$) is added that indicates whether it was a forest cover increase or decrease. 
A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have a linear or non-linear relationship with $\Delta Qf \%$. This means that a smooth function $s()$ is applied to the $Z_i$ variables. For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty which means the terms can be shrunk to 0 if not significant [@wood2006].

@@

<!-- analysis -->
## description of the data

The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017). The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^.

```{r, echo = F, warning = F, message=F, eval =F}
png("DataExploration.png", width = 2500, height = 1800, res = 300)
All_data %>%
  pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm, E0), 
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() +
  facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
    c("Area_km2" = "Area (km2)",
      "DeltaF_perc" = "Delta Forest (%)",
      "DeltaQf_perc" = "Delta Qf (%)",
      "E0" = "Annual E0 (mm)",
      "Pa_mm" = "Annual Rainfall (mm)"))) +   
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5)),
        axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()
```

@@

```{r, echo = F}
All_data2 <- All_data %>%
  mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
                              "Decrease", "Increase"),
         DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
                                  -1*DeltaF_perc,
                                  DeltaF_perc))
```

```{r, echo = F, warning = F, message=F, eval =F}
# plot to show the difference
png("Increase_decrease.png", width = 2500, height = 1800, res = 300)
All_data2 %>%
  filter(is.na(DeltaF_perc_pos) == F) %>%
  ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
  geom_point() +  
  facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
  theme_bw() + 
    theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5))) + 
  scale_colour_manual(name = "Forest change direction", 
                      values = c("Decrease" = "red", "Increase" = "blue")) +
  xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
dev.off()

@@

@@

```{r, echo =F}
pander(round(summary(model4_dryness)$p.table,2), caption = "(\\#tab:out-model4) Results of the model including the dryness index")
```
<!-- running PCA, this is used later -->

@@

The major hypothesis to test is the effect of area on the change in flow, following the analysis by @zhang2017 and @filoso2017. Given the highly skewed distribution of the catchment areas (Figure \@ref(fig:datagraphs)), a log base 10 transformation was applied to the variable *Area (km^2^)*. The effect of Dryness remains in the model, given that there could be interactions that are worth exploring.

@@

```{r modelArea_int, echo = F, eval =T}
model_Area_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
                        Dryness*log10(Area_km2), data = All_data2)
#summary(model_Area_b)
#gam.check(model_Area)
```

@@

This model explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`. This indicates that the categorical variables explain a limited amount of the variance. However, interesting to note from Table \@ref(tab:msix-linear) that several of the assessment methods are significant. In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant ($p < 0.05$). 
In this case, *Area (km^2)* is no longer a significant predictor, the reasons for this will be discussed later.

@@

<!--filter length < 60 -->
```{r model7_nolongLength, echo = F}
NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4, From > 1920)
model7 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
                      s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + s(From, bs = "ts") +
                    Precip_data_type +  Assessment_technique + Forest_type +
                    Hydrological_regime, data = NoLong_Data)
#summary(model7)
```
```{r, echo = F}
pander(round(summary(model7)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years, Dryness > 4 and studies commencing after 1920")
```
```{r visualisesmooths7,echo =F, eval = F}
#require(visreg)
png("model7_smooths.png", width = 2500, height = 1800, res = 300)
#par(mfrow=c(2,2))
D <- visreg(model7, "Dryness", gg = T, plot = T) + theme_bw()
A <- visreg(model7, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
L <- visreg(model7, "length", gg = T, plot = T) + theme_bw()
From <- visreg(model7, "From", gg = T, plot = T) + theme_bw()
(D + A) / (L + From) 
# visreg(model6_all,"Dryness",gg=T) + theme_bw()
# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
#par(mfrow=c(1,1))
dev.off()
```
```{r smoothsmodel7, fig.cap="Visualisation of the smooth variables in the model with reduced data for dryness and length", out.width="90%", echo=F}

@@

The flexible nature of the splines means that the length variable captures some substantial variation in the data, but it is unclear what exactly is captured. The shape of the conditional response (Figure \@ref(fig:smoothsmodel6)) also does not reflect the type of response highlighted in @filoso2017 and @jackson2005. One reason could be the few data points with very long data series, and very old studies (before 1920 essentially), and highly variable responses (Figure \@ref(fig:smoothsmodel6)). Therefore it can be important to investigate what removing these factors has on the overall model and the significance of the variables. The next model therefore removes the following data: *Dryness* > 4, *length* > 60 years, *From* > 1920. This result in a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments.
The model has more explaining power with an adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`.  This results in Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover on streamflow (Figure \@ref(fig:smoothsmodel7)). However the other 3 variables are no longer significant due to the scatter in the data, but *length* and *From* show some strong local non-linear behaviour.

@@

Stronger statements about the trends in the change in flow cannot be made until more data or better data becomes available in this area. Furthermore, the current study analyses a large global dataset. This analysis does not exlude more local and regional effects that cannot be identified in the global data.

>>>>

---
title: Factors determining how catchments respond to forest cover change. Re-analysing global data sets.
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: DARE, The University of Sydney
    footnote: 1
  - name: Eliana Nervi
    email: eliananervi@gmail.com
    affiliation: INIA
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
    affiliation: IMFIA
address:
  - code: DARE
    address: ARC Training Centre Data Analytics for Resources and the Environment
  - code: The University of Sydney
    address: School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia
  - code: INIA
    address: Project Manager, FPTA 357, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
  - code: IMFIA
    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow. Deforestation causes a 32\% greater change in flow compared to reforestation. Area of the catchment only affects the change in streamflow after log transformation, given the wide variety in the data from small scale paired watershed studies. Smaller studies dominate the database with 42\% of the data < 1 km2 and 65\% of the data < 10 km2. Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies. Despite these findings, overall explained variance (38\%) is low due the quality of the inputs and additional unknown confounding factors.
  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow. However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation. The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow? Generalised additive models were used to run flexible regressions including multiple variables. 
journal: "Journal of Hydrology"
date: "`r Sys.Date()`"
bibliography: forest_and_water.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
editor_options: 
  chunk_output_type: inline
---

@@

There has been an long and on-going discussion in the hydrological literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017]. The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa. The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover. However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchments after deforestation.

@@

In addition, additional variables added were the latitude and longitude for the center of the catchment as an approximation of its spatial location. Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers. For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.

@@

Several additional data points from catchment studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e. Table 6 in @zhang2011).
We also removed one data point from the analysis, which corresponds to catchment #1 (Amazon) in @zhang2017. This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded. Furthermore, the change in flow for catchment #76 was corrected from 600% to 157% after review of the original publication [@baker1984]. Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data.

@@

## Statistical modelling
```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```
<!----
The starting point consists of the two tables in the supplementary data from Zhang et al. (2017), which means there is one table for small watersheds (< 1000 km2) and one table for large watersheds (> 1000 km2). We might combine these tables as we can easily add a variable indicating whether the watershed originates from the small or large table.
--->
```{r, echo = F, message = F}
require(tidyverse)
require(lubridate)
require(mgcv)
require(visreg)
require(patchwork)
suppressWarnings(require(knitr))
suppressWarnings(require(bookdown))
suppressWarnings(require(pander))

@@

<!-- read in the ET data -->

```{r, echo = F, warning = F, message=F}
PET_large <- read_csv("../../data/largeET.csv")
PET_small <- read_csv("../../data/smallET.csv")

@@

# 
PET_all <- bind_rows(PET_large, PET_small)
# There are 7 missing E0 values
# > PET_all$`Watershed #`[is.na(PET_all$E0)==T]
# [1]  38 278 269 243 236 299 250
```

@@

<!--now combine with E0 values -->
```{r, echo = F}
All_data <- left_join(All_data ,PET_all %>% select(-c(Longitude, Latitude)), by = "Watershed #")
All_data <- All_data %>%
  mutate(E0 = ifelse(is.na(E0.x)==T, E0.y, E0.x)) %>%
  select(-c(E0.x,E0.y)) #%>%
  #rename(Latitude = Latitude.x,
#         Longitude = Longitude.x)

@@

Here $X_i$ are factorial variables, while $Z_i$ are continuous variables. The model assumes no direct interactions and all variables are additive. The changes in forest cover contain both positive (forestation) and negative values (deforestation). In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect if removing forest cover was the same as an equivalent reforestation. However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover. Therefore all the change in forest cover data is converted to positive values, and an additional column ($sign_{forest cover}$) is added that indicates whether it was a forest cover increase or decrease.

A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have a linear or non-linear relationship with $\Delta Qf \%$. This means that a smooth function $s()$ is applied to the $Z_i$ variables. For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty which means the terms can be shrunk to 0 if not significant [@wood2006].

@@

<!-- analysis -->
## Description of the data

The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017). The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and *Area km^2*. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^ (Figure 1).

```{r, echo = F, warning = F, message=F, eval =F}
png("DataExploration.png", width = 2500, height = 1800, res = 300)
All_data %>%
  pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm, E0), 
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() +
  facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
    c("Area_km2" = "Area (km2)",
      "DeltaF_perc" = "Delta Forest (%)",
      "DeltaQf_perc" = "Delta Qf (%)",
      "E0" = "Annual Potential E (mm)",
      "Pa_mm" = "Annual Rainfall (mm)"))) +   
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5)),
        axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()
```

@@

```{r, echo = F}
All_data2 <- All_data %>%
  mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
                              "decrease", "increase"),
         DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
                                  -1*DeltaF_perc,
                                  DeltaF_perc))
```

In addition, this shows that for the data related to forest decreases, there is almost always a positive flow change (Figure 2). In other words, flow almost always increased. However, for increases in forest cover, this is not the case, and flow can both increase and decrease. However in both cases the variability in the reported change in flow increases with the increase in forest cover change.

```{r, echo = F, warning = F, message=F, eval =F}
# plot to show the difference
png("Increase_decrease.png", width = 2500, height = 1800, res = 300)
All_data2 %>%
  ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
  geom_point() +  
  facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
  theme_bw() + 
    theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5))) + 
  scale_colour_manual(name = "Forest change direction", 
                      values = c("Decrease" = "red","Increase" = "blue")) +
  xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
dev.off()

@@

<!--\newpage-->

@@

```{r, echo =F}
pander(round(summary(model4_dryness)$p.table,2), caption = "(\\#tab:out-model4) Results of the model including the dryness index")
```

<!-- running PCA, this is used later -->

@@

The major hypothesis to test is the effect of area on the change in flow, following the analysis by @zhang2017 and @filoso2017. Given the highly skewed distribution of the catchment areas (Figure \@ref(fig:datagraphs)), a log base 10 transformation was applied to the variable *Area (km^2^)*.

@@

```{r modelArea_b, echo = F, eval =T}
#Zhang_all2 <- Zhang_all2 %>%
#  mutate(length = To - From)

model_Area_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
                        Dryness*log10(Area_km2), data = All_data2)
#summary(model_Area_b)
#gam.check(model_Area)
```

@@

This model explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`. This indicates that the categorical variables explain a limited amount of the variance. However, interesting to note from Table \@ref(tab:msix-linear) that several of the assessment methods are significant. In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant ($p < 0.05$). 
In this case, *log10(Area (km^2^))* is no longer a significant predictor, the reasons for this will be discussed later.

@@

<!--filter length < 60 -->
```{r model7_nolongLength, echo = F}

NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4, From > 1920)

model7 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
                      s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + s(From, bs = "ts") +
                    Precip_data_type +  Assessment_technique + Forest_type +
                    Hydrological_regime, data = NoLong_Data)
#summary(model7)
```

```{r, echo = F}
pander(round(summary(model7)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years, Dryness > 4 and studies commencing after 1920")
```

```{r visualisesmooths7,echo =F, eval = F}
#require(visreg)
png("model7_smooths.png", width = 2500, height = 1800, res = 300)
#par(mfrow=c(2,2))
D <- visreg(model7, "Dryness", gg = T, plot = T) + theme_bw()
A <- visreg(model7, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
L <- visreg(model7, "length", gg = T, plot = T) + theme_bw()
From <- visreg(model7, "From", gg = T, plot = T) + theme_bw()
(D + A) / (L + From) 
# visreg(model6_all,"Dryness",gg=T) + theme_bw()
# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
#par(mfrow=c(1,1))
dev.off()
```

```{r smoothsmodel7, fig.cap="Visualisation of the smooth variables in the model wiht reduced data for dryness and length", out.width="90%", echo=F}

@@

<!--\newpage-->

The flexible nature of the splines means that the length variable captures some substantial variation in the data, but it is unclear what exactly is captured. The shape of the conditional response (Figure \@ref(fig:smoothsmodel6)) also does not reflect the type of response highlighted in @filoso2017 and @jackson2005. One reason could be the few data points with very long data series, and very old studies (before 1920 essentially), and highly variable responses (Figure \@ref(fig:smoothsmodel6)). Therefore it can be important to investigate what removing these factors has on the overall model and the significance of the variables. The next model therefore removes the following data: *Dryness* > 4, *length* > 60 years, *From* > 1920. This result in a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments.

The model has more explaining power with an adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`.  This results in Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover on streamflow (Figure \@ref(fig:smoothsmodel7)). However the other 3 variables are no longer significant due to the scatter in the data, but *length* and *From* show some strong local non-linear behaviour.

@@

Stronger statements about the trends in the change in flow cannot be made until more data or better data becomes available in this area. Furthermore, the current study analyses a large global dataset. This analysis does not exclude more local and regional effects that cannot be identified in the global data.
